Predicting the Ending Notes of Bach Chorales

### 2018-05-08 Paper.tex
Add index to confusion matricies

Johann Sebastian Bach, as a product of the Baroque period, followed a set of composition guidelines which gave the Common Practice Period its name. These guidelines dealt with methods of resolving dissonance, writing melodies, and more. Despite decades and centuries of experimentation since Bach's death in 1750, many of these common practices can still be seen in Western music. Thus, his work makes for an excellent starting point for predictive modeling.

\section{Computer-Generated Music}
There have been a number of attempts to create algorithmic composition software, a program which can compose music with little human interaction or involvement; one of the first examples dating to the 1950's when Lejaren Hiller and Leonard Isaacson produced {\it The Illiac Suite for String Quartet} \cite{computers}. There are two varieties of algorithmic composition: top-down and bottom-up; an implementation of the bottom-up approach is covered in detail later.\\

@@

\subsection{A Bottom-Up Approach: Splicing Systems}
In their attempt to create an bottom-up algorithmic composition model, De Felice et. al. use a technique originally conceived for DNA analysis called a {\it splicing system} to teach a computer to compose original music in the style of Bach \cite{computer_composer}. After converting the chorales into their own tonality degree representation (a notation system that records the voice, pitch, octave, and major/minor tonality of each note), each word is passed into the splicing system. It is in this splicing system that the harmonic function (i.e. the chord progression) and the melodic function (how each note relates to the chord) are developed from the input information. The final model uses a combination of predefined rules of classical music and statistical information extracted from existing data in its compositions.\\

@@

\section{Data}
The data set used to test the models is a collection of one hundred Bach chorales, provided by the University of California Irvine's Machine Learning Repository \cite{data_set}. Each chorale contains only the melody line from each chorale. Each note in the melody has six different information fields. The starting time of the note is measured in 16th notes from the start of the piece. The pitch of the note is provided in the MIDI tuning standard, with the note C4 given a value of 60, C\#4 given a value of 61, etc. The pitches are restricted to values between 60 and 75, by nature of the notes present in the chorales. The duration of each note is given, also measured in 16th notes. The key signature is presented as values between -4 and +4. The numbering relates to the number of sharps or flats are present in the key signature: positive for sharps, negative for flats. The time signature is again measured in the number of 16th notes, this time in the number of 16th notes per measure. Finally, a Boolean flag is set to true if the note is underneath a fermata. Fermatas generally indicate the end of a musical phrase or chord progression, and instruct the performer to hold the note for as long as deemed necessary.

\section{Methods}
Three separate models were created for this experiment: a naive baseline model, a human-made model with predefined rules, and a computer-generated model built on classification trees. In addition, three variations of the computer model were built. Two used the same structure as the original, but with smaller subsets of the data; the third variation used a k-nearest-neighbors tree.

@@

\subsection{Human Model}
To predict the final note, the smart model used the two preceding notes as reference. Looking at the key signature and the pitch, the model performs modulo operations on the MIDI pitch to find the relative position in the key, with values between 0 and 11. For example, the tonic becomes 0, the major third becomes 4, and the perfect fifth becomes 7. With these values calculated, the model uses a defined set of patterns to predict the melody endings. For the sake of simplicity, only notes in the major triad were used in this model's predictions.\\

@@

This model predicted the endings of a total of 56 chorales correctly. Despite only a minor increase in correct pitches, the MSE decreased dramatically to 4.32. While the number of underestimations was reduced by around 60\%, there was a 50\% rise in the number of overestimations. This switch may be caused by a number of possibilities. Since the model only selected notes within the major triad, any chorale which ended on a minor third would be missed by the model. There are likely some patterns which are not considered by the algorithm, where the two penultimate notes have two possible endings. In these cases, only the more prevalent one was incorporated into the algorithm. Finally, any chorale which uses the last two notes as part of a key change will not be accurately predicted, since this model uses the relative position within the current key.

@@

Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy. The confusion matrix for the computer model has one less row and column than both the naive model and 3 fewer rows and columns than the human model. The pitches for the notes were given in a range between 60 and 75 in the MIDI scale, but this range applies to the entire melody line, not the endings. While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.\\

>>>>

Johann Sebastian Bach, as a product of the Baroque period (1600-1750), followed a set of composition guidelines which gave the Common Practice Period its name. These guidelines dealt with methods of resolving dissonance, writing melodies, and more. Despite decades and centuries of experimentation since Bach's death in 1750, many of these common practices can still be seen in Western music. Thus, his work makes for an excellent starting point for predictive modeling.

\section{Computer-Generated Music}
There have been a number of attempts to create algorithmic composition software, a program which can compose music with little human interaction or involvement; one of the first examples dating to the 1950's when Lejaren Hiller and Leonard Isaacson produced {\it The Illiac Suite for String Quartet} \cite{computers}. There are two varieties of algorithmic composition: top-down and bottom-up.\\

@@

\subsection{A Bottom-Up Approach: Splicing Systems}
In their attempt to create a bottom-up algorithmic composition model, De Felice et. al. use a technique originally conceived for DNA analysis called a {\it splicing system} to teach a computer to compose original music in the style of Bach \cite{computer_composer}. After converting the chorales into their own tonality degree representation (a notation system that records the voice, pitch, octave, and major/minor tonality of each note), each word is passed into the splicing system. It is in this splicing system that the harmonic function (i.e. the chord progression) and the melodic function (how each note relates to the chord) are developed from the input information. The final model uses a combination of predefined rules of classical music and statistical information extracted from existing data in its compositions.\\

@@

\section{Data}
The data set used to test the models is a collection of 100 Bach chorales, provided by the University of California Irvine's Machine Learning Repository \cite{data_set}. Each chorale contains only the melody line from each chorale. Each note in the melody has six different information fields. The starting time of the note is measured in 16th notes from the start of the piece. The pitch of the note is provided in the MIDI tuning standard, with the note C4 given a value of 60, C$^{\#}$4 (C sharp) given a value of 61, etc. The pitches are restricted to values between 60 and 75, by nature of the notes present in the chorales. The duration of each note is given, also measured in 16th notes. The key signature is presented as values between -4 and +4. The numbering relates to the number of sharps or flats are present in the key signature: positive for sharps, negative for flats. The time signature is again measured in the number of 16th notes, this time in the number of 16th notes per measure. Finally, a Boolean flag is set to true if the note is underneath a fermata. Fermatas generally indicate the end of a musical phrase or chord progression, and instruct the performer to hold the note for as long as deemed necessary.

\section{Methods}
Three separate models were created for this experiment: a naive baseline model, a human-made model with predefined rules, and a computer-generated model built on classification trees. In addition, three variations of the computer model were built. Two used the same structure of classification trees as the main model, but with smaller subsets of the data; the third variation used a k-nearest-neighbors tree.

@@

\subsection{Human Model}
To predict the final note, the smart model used the two preceding notes as reference. Looking at the key signature and the pitch, the model performs modulo operations on the MIDI pitch to find the relative position in the key, with values between 0 and 11. For example, the tonic becomes 0, the major third becomes 4, and the perfect fifth becomes 7. With these values calculated, the model uses a defined set of patterns to predict the melody endings. For the sake of simplicity, only notes in the major triad (tonic, major third, perfect fifth) were used in this model's predictions.\\

@@

This model predicted the endings of a total of 56 chorales correctly. Despite only a minor increase in correct pitches, the MSE decreased dramatically to 4.32. While the number of underestimations was reduced by around 60\%, there was a 50\% rise in the number of overestimations. This switch may be caused by a number of possibilities. Since the model only selected notes within the major triad, any chorale which ended on a minor third would be missed by the model. There are likely some patterns which are not considered by the algorithm, cases where the same pattern of two notes has two possible endings. In these cases, only the more prevalent one was incorporated into the algorithm. Finally, any chorale which uses the last two notes as part of a key change will not be accurately predicted, since this model uses the relative position within the current key.

@@

Although the human and computer models performed at similar levels, one noticeable difference between the two is accuracy. The confusion matrix for the computer model has one less row and column than both the naive model and 3 fewer rows and columns than the human model. The pitches for the notes were given in a range between 60 (C4) and 75 (E$^b$5) in the MIDI scale, but this range applies to the entire melody line, not the endings. While the human model did not assume any distribution of pitches in the range, the computer model determined that the ending notes are clustered in the middle of the range, thus explaining the shrinkage of the confusion matrix.\\
